### 2019-06-20 1-Preamble/4-Acknowledgement/acknowledgement.tex
Huge proofreading!

Thanks to my wonderful colleagues during these three years: Rémi$^*$, Pascal$^*$, Yves, Jacques, Ali, Haïfa, Marwa, Navik, Quentin$^*$, Vincent$^*$, Rami$^*$, Muhammad, Cristo$^*$, Adrien$^*$, Esteban$^*$, Ali, Éloïse$^*$, Nabil$^*$, Bastien$^*$, Corentin$^*$, Georgios, Morgane$^*$ in Rennes ; and Ronan, Odalric-Ambrym, Philippe, Olivier, Florian, Xuedong, Guillaume, Julien, Mathieu, Matteo, Mahsa, Nicolas, Édouard in Lille.
I want to thank especially the ``coffee friends$^*$'' in Rennes, for the daily interesting discussions, which where sometimes the highlights of our life at the office.
%
Thanks to everybody who played \emph{futsal} on Wednesday lunch, I still have to progress immensely before I could pretend to be useful, but it's been a lot of fun to play with you guys!

>>>>

Thanks to my wonderful colleagues during these three years: Rémi$^*$, Pascal$^*$, Yves, Jacques, Ali, Haïfa, Marwa, Navik, Quentin$^*$, Vincent$^*$, Rami$^*$, Muhammad, Cristo$^*$, Adrien$^*$, Esteban$^*$, Ali, Éloïse$^*$, Nabil$^*$, Bastien$^*$, Corentin$^*$, Georgios, Morgane$^*$ in Rennes ; and Alessandro, Ronan, Odalric-Ambrym, Philippe, Olivier, Florian, Xuedong, Guillaume, Julien, Mathieu, Matteo, Mahsa, Nicolas, Édouard in Lille.
I want to thank especially the ``coffee friends$^*$'' in Rennes, for the daily interesting discussions, the best moments of our life at the office.
%
Thanks to everybody who played \emph{futsal} on Wednesday lunch, I still have to progress immensely before I could pretend to be useful, but it's been a lot of fun to play with you guys!

### 2019-06-20 1-Preamble/5-Abstract/abstract.tex
Huge proofreading!

% \newpage
% ----------------------------------------------------------------------------
\begin{center}
	\textbf{{\LARGE Abstract}}
\end{center}
% \vspace*{1cm}
% \vspace*{-5pt}
%
% \begin{small} % XXX
% Chapter 1
In this PhD thesis, we study wireless networks and reconfigurable end-devices that can access large-scale Cognitive Radio networks, in unlicensed bands and without central control.
We focus on Internet of Things networks (IoT), with the objective of extending the devices' battery life, by equipping them with low-cost but efficient machine learning algorithms, to let them automatically improve the efficiency of their wireless communications.
%  (Chapter~\ref{chapter:1})
% Chapter 4
We propose different models of IoT networks, and we show empirically on both numerical simulations and real-world validation the possible gain of our methods, that use Reinforcement Learning.
%  (Chapter~\ref{chapter:4})
% Chapter 2
The different network access problems are modeled as Multi-Armed Bandits (MAB), but we found that analyzing the realistic models was intractable,
% ,  Chapter~\ref{chapter:2}
because proving the convergence of many end-devices playing a collaborative game without communication no coordination is hard, when end-devices all follow random active pattern.
% Indeed, even the number of active devices at any time is highly unpredictable, making the environment not stationary and quickly evolving.
% Chapter 5 and 6
The rest of this manuscript thus studies two restricted models, first multi-players bandits in stationary problems, then non-stationary single-player bandits.
%  (Chapter~\ref{chapter:5})
%  (Chapter~\ref{chapter:6})
% Chapter 3
We also detail another contribution, SMPyBandits, an open-source Python library for numerical MAB simulations, covering all the studied models and more.
%  (Chapter~\ref{chapter:3})
% \end{small} % XXX

>>>>

% \newpage
% ----------------------------------------------------------------------------
\begin{center}
	\textbf{{\LARGE Abstract}}
\end{center}
% \vspace*{1cm}
% \vspace*{-5pt}
%
% \begin{small} % XXX
% Chapter 1
In this PhD thesis, we study wireless networks and reconfigurable end-devices that can access large-scale Cognitive Radio networks, in unlicensed bands and without central control.
We focus on Internet of Things networks (IoT), with the objective of extending the devices' battery life, by equipping them with low-cost but efficient machine learning algorithms, to let them automatically improve the efficiency of their wireless communications.
%  (Chapter~\ref{chapter:1})
% Chapter 4
We propose different models of IoT networks, and we show empirically on both numerical simulations and real-world validation the possible gain of our methods, that use Reinforcement Learning.
%  (Chapter~\ref{chapter:4})
% Chapter 2
The different network access problems are modeled as Multi-Armed Bandits (MAB), but we found that analyzing the realistic models was intractable,
% ,  Chapter~\ref{chapter:2}
because proving the convergence of many end-devices playing a collaborative game without communication nor coordination is hard, when end-devices all follow random active pattern.
% Indeed, even the number of active devices at any time is highly unpredictable, making the environment not stationary and quickly evolving.
% Chapter 5 and 6
The rest of this manuscript thus studies two restricted models, first multi-players bandits in stationary problems, then non-stationary single-player bandits.
%  (Chapter~\ref{chapter:5})
%  (Chapter~\ref{chapter:6})
% Chapter 3
We also detail another contribution, SMPyBandits, an open-source Python library for numerical MAB simulations, covering all the studied models and more.
%  (Chapter~\ref{chapter:3})
% \end{small} % XXX

### 2019-06-20 2-Chapters/2-Chapter/chapter2.tex
Huge proofreading!

The difficulty of this decision-making game comes from balancing the trade-off between \emph{exploration}, because the player has to observe as much as possible all the arms to get more knowledge about the distributions of their rewards, and \emph{exploitation}, because she also has to select as much as possible the best arm.
The MAB model is a famous example of a \textbf{reinforcement learning} model, where the decision-making process has to adapt to an unknown environment using noisy observations and a discrete time, alternating between decisions and observations.
We illustrate this cycle in Figure~\ref{fig:2:ReinforcementLearningCycleMABmodel} below.
For more details on reinforcement learning, in particular about more generic models, we refer to the famous book by Sutton and Barto \cite{SuttonBarto2018}.

@@

Given a measure $\lambda$, that is usually the natural Lebesgue measure on $\R$, an exponential family of probability distributions is defined as the distributions whose density (relative to $\lambda$) can be written as
% \begin{equation}\label{eq:2:exponentialFamily}
$ \Pr_{\eta}(x | \lambda) = h(x) \exp \left( \eta x - A(\eta) \right)$,
% \end{equation}
for a parameter vector $\eta$ (the canonical parameter),
% and a sufficient statistic that we restrict to be $T(x)=x$,
and a function $h$.
The cumulant function $A(\eta)$ is entirely determined by the $\eta$ and $h$,
as $A(\eta) = \log \left( \int h(x) \exp(\eta x) \lambda(\mathrm{d} x) \right)$.
%
The natural parameter space is the set of values of $\eta$ such that this integral $A(\eta)$ is finite,
and usually the literature focusses on regular and minimal exponential families (when the natural parameter space is an non-empty open set in $\R$).

@@

In all this section, we restrict to stationary stochastic problems with $K\geq2$ arms.
We do not give proofs of the following theorems, as they can be found in the historical papers, and simpler proofs are given in recent references, such as \cite{Bubeck12}, \cite{LattimoreBanditAlgorithmsBook}, or Chapter~2 in \cite{Slivkins2019} for instance.
%
Let $\cI$ denote the set of all problem instances, with $K \geq 2$ arms.
We assume the rewards lie in $[0,1]$,
to present the lower-bounds, we prefer to focus on Bernoulli distributions, for simplicity.

To specify the dependency on an instance $I\in\cI$, we denote the regret of $\cA$ on instance $I$ and horizon $T$ by $R_T^{\cA}(I)$.
We also index the Landau notations $o(\dots)$ and $\bigO{\dots}$ by $I$,
for instance $R_T^{\cA}(I) = o_I(\log(T))$ means $R_T^{\cA}(I) = \smallO{c_I \log(T)}$
where the ``constant'' $c_I$ can depend on the problem instance $I$ (\eg, on its means and on $K$) but \emph{not} on the time horizon $T$.

@@

\begin{table}[!h]
    \begin{small}  % WARNING
        \centering
        \begin{tabular}{c|cc|cc}
        \textbf{Algorithms} & $\sum\limits_{t=1}^{T=100} r(t)$ & $R_{T=100}^{\cA}$ & $\sum\limits_{t=1}^{T=10000} r(t)$ & $R_{T=10000}^{\cA}$ \\
            \hline
            Pure exploration
                & $46$ & $20.9$
                & $5171$ & $1886.4$ \\
            \textbf{Pure exploitation}
                & $\mathbf{65}$ & $\mathbf{1.15}$
                & $\mathbf{6997}$ & $\mathbf{0.95}$ \\
            \hline
            $\varepsilon$-greedy
                & $50$ & $20.9$
                & $6657$ & $382.2$ \\
            Explore-then-Commit
                & $49$ & $15$
                & $5766$ & $1199.8$ \\
            \hline
            $\UCB_1$
                & $63$ & $10.3$
                & $6920$ & $78.4$ \\
            \klUCB{}
                & $56$ & $12.4$
                & $6922$ & $70.2$ \\
            \hline
            Thompson sampling
                & $51$ & $12.2$
                & $6949$ & $44.2$ \\
        \end{tabular}
        \caption{Cumulated rewards and regret, for horizons $T=100$ and $T=10000$, \emph{for only one run of the simulation}, for different algorithms.}
        \label{table:2:oneShotResult}
    \end{small}  % WARNING
\end{table}

@@

\begin{table}[ht]
    \begin{small}  % WARNING
        \centering
        \begin{tabular}{c|cc|cc}
        \textbf{Algorithms} & $\frac{1}{N} \sum\limits_{j=1}^N \sum\limits_{t=1}^{T=100} r^{(j)}_{A(t)}(t)$ & $\frac{1}{N} \sum\limits_{j=1}^N R^{\cA,(j)}_{T=100}$ & $\frac{1}{N} \sum\limits_{j=1}^N \sum\limits_{t=1}^{T=10000} r^{(j)}_{A(t)}(t)$ & $\frac{1}{N} \sum\limits_{j=1}^N R^{\cA,(j)}_{T=10000}$ \\
            \hline
            Pure exploration
                & $51$ & $19$
                & $5099$ & $1900$ \\
            Pure exploitation
                & $62$ & $8.3$
                & $6349$ & $653$ \\
            \hline
            $\varepsilon$-greedy
                & $51$ & $18.7$
                & $6460$ & $542$ \\
            Explore-then-Commit
                & $57$ & $12.9$
                & $5795$ & $1207$ \\
            \hline
            $\UCB_1$
                & $\mathbf{60}$ & $\mathbf{9.8}$
                & $6921$ & $79$ \\
            \klUCB{}
                & $61$ & $9.3$
                & $6927$ & $72$ \\
            \hline
            Thompson sampling
                & $\mathbf{60}$ & $\mathbf{9.6}$
                & $\mathbf{6951}$ & $\mathbf{49}$ \\
        \end{tabular}
        \caption{Cumulated rewards and regret, for horizons $T=100$ and $T=10000$, \emph{averaged over $N=1000$ independent simulations} ($j=1,\dots,N$), for different algorithms.}
        \label{table:2:meanResults}
    \end{small}  % WARNING
\end{table}

>>>>

The difficulty of this decision-making game comes from balancing the trade-off between \emph{exploration}, because the player has to observe as much as possible all the arms to get more knowledge about the distributions of their rewards, and \emph{exploitation}, because she also has to select as much as possible the best arm.
The MAB model is a famous example of a \textbf{reinforcement learning} model, where the decision-making process has to adapt to an unknown environment using noisy observations and a discrete time, alternating between decisions and observations.
We illustrate this cycle in Figure~\ref{fig:2:ReinforcementLearningCycleMABmodel} below.
For more details on reinforcement learning, in particular about more generic models, we refer to the famous book \cite{SuttonBarto2018}.

@@

Given a measure $\lambda$, that is usually the natural Lebesgue measure on $\R$, an exponential family of probability distributions is defined as the distributions whose density (relative to $\lambda$) can be written as
% \begin{equation}\label{eq:2:exponentialFamily}
$ \Pr_{\eta}(x | \lambda) = h(x) \exp \left( \eta x - A(\eta) \right)$,
% \end{equation}
for a parameter vector $\eta$ (the canonical parameter),
% and a sufficient statistic that we restrict to be $T(x)=x$,
and a function $h$.
The cumulant function $A(\eta)$ is entirely determined by $\eta$ and $h$,
as $A(\eta) = \log \left( \int h(x) \exp(\eta x) \lambda(\mathrm{d} x) \right)$.
%
The natural parameter space is the set of values of $\eta$ such that this integral $A(\eta)$ is finite,
and usually the literature focusses on regular and minimal exponential families (when the natural parameter space is an non-empty open set in $\R$).

@@

In all this section, we restrict to stationary stochastic problems with $K\geq2$ arms.
We do not give proofs of the following theorems, as they can be found in the historical papers, and simpler proofs are given in recent references, such as \cite{Bubeck12}, \cite{LattimoreBanditAlgorithmsBook}, or Chapter~2 in \cite{Slivkins2019} for instance.
%
Let $\cI$ denote the set of all problem instances, with $K \geq 2$ arms.
We assume the rewards lie in $[0,1]$,
to present the lower-bounds, we prefer to focus on Bernoulli distributions, for simplicity.
%
To specify the dependency on an instance $I\in\cI$, we denote the regret of $\cA$ on instance $I$ and horizon $T$ by $R_T^{\cA}(I)$.
We also index the Landau notations $o(\dots)$ and $\bigO{\dots}$ by $I$,
for instance $R_T^{\cA}(I) = o_I(\log(T))$ means $R_T^{\cA}(I) = \smallO{c_I \log(T)}$
where the ``constant'' $c_I$ can depend on the problem instance $I$ (\eg, on its means and on $K$) but \emph{not} on the time horizon $T$.

@@

\begin{table}[!h]
    \begin{small}  % WARNING
        \centering
        \begin{tabular}{c|cc|cc}
        \textbf{Algorithms}
        & rewards
        & regret for $T=100$
        & rewards
        & regret for $T=10000$ \\
        % & $\sum\limits_{t=1}^{T=100} r(t)$ & $R_{T=100}^{\cA}$ & $\sum\limits_{t=1}^{T=10000} r(t)$ & $R_{T=10000}^{\cA}$ \\
        % & $\sum\limits_{t=1}^{T=100} r(t)$ & $R_{T=100}^{\cA}$ & $\sum\limits_{t=1}^{T=10000} r(t)$ & $R_{T=10000}^{\cA}$ \\
            \hline
            Pure exploration
                & $46$ & $20.9$
                & $5171$ & $1886.4$ \\
            \textbf{Pure exploitation}
                & $\mathbf{65}$ & $\mathbf{1.15}$
                & $\mathbf{6997}$ & $\mathbf{0.95}$ \\
            \hline
            $\varepsilon$-greedy
                & $50$ & $20.9$
                & $6657$ & $382.2$ \\
            Explore-then-Commit
                & $49$ & $15$
                & $5766$ & $1199.8$ \\
            \hline
            $\UCB_1$
                & $63$ & $10.3$
                & $6920$ & $78.4$ \\
            \klUCB{}
                & $56$ & $12.4$
                & $6922$ & $70.2$ \\
            \hline
            Thompson sampling
                & $51$ & $12.2$
                & $6949$ & $44.2$ \\
        \end{tabular}
        \caption{Cumulated rewards and regret, for horizons $T=100$ and $T=10000$, \emph{for only one run of the simulation}, for different algorithms.}
        \label{table:2:oneShotResult}
    \end{small}  % WARNING
\end{table}

@@

\begin{table}[ht]
    \begin{small}  % WARNING
        \centering
        \begin{tabular}{c|cc|cc}
        \textbf{Algorithms}
        & rewards
        & regret for $T=100$
        & rewards
        & regret for $T=10000$ \\
        % & $\frac{1}{N} \sum\limits_{j=1}^N \sum\limits_{t=1}^{T=100} r^{(j)}_{A(t)}(t)$ & $\frac{1}{N} \sum\limits_{j=1}^N R^{\cA,(j)}_{T=100}$ & $\frac{1}{N} \sum\limits_{j=1}^N \sum\limits_{t=1}^{T=10000} r^{(j)}_{A(t)}(t)$ & $\frac{1}{N} \sum\limits_{j=1}^N R^{\cA,(j)}_{T=10000}$ \\
            \hline
            Pure exploration
                & $51$ & $19$
                & $5099$ & $1900$ \\
            Pure exploitation
                & $62$ & $8.3$
                & $6349$ & $653$ \\
            \hline
            $\varepsilon$-greedy
                & $51$ & $18.7$
                & $6460$ & $542$ \\
            Explore-then-Commit
                & $57$ & $12.9$
                & $5795$ & $1207$ \\
            \hline
            $\UCB_1$
                & $\mathbf{60}$ & $\mathbf{9.8}$
                & $6921$ & $79$ \\
            \klUCB{}
                & $61$ & $9.3$
                & $6927$ & $72$ \\
            \hline
            Thompson sampling
                & $\mathbf{60}$ & $\mathbf{9.6}$
                & $\mathbf{6951}$ & $\mathbf{49}$ \\
        \end{tabular}
        \caption{Cumulated rewards and regret, for horizons $T=100$ and $T=10000$, \emph{averaged over $N=1000$ independent simulations} ($j=1,\dots,N$), for different algorithms.}
        \label{table:2:meanResults}
    \end{small}  % WARNING
\end{table}
